---
output: github_document
---

# historian

```{r}
library(tidyverse)
```

## Problem

We want to keep track of the state of a table at different points in time.
The table has a primary key `id` and a column `x` that we want to keep track of.
In this example, `V1` is the initial state of the table, `V2` is the state of the table after adding a row, `V3` is the state of the table after modifying a row, and `V4` is the state of the table after deleting a row.

```{r}
V1 <- tibble(id = 1L, x = letters[1])
V1

# Adding a row
V2 <- tibble(id = 1:2, x = letters[1:2])
V2

# Modifying a row
V3 <- tibble(id = 1:2, x = letters[3:2])
V3

# Deleting a row
V4 <- tibble(id = 2L, x = letters[2])
V4
```

## History table

At each point in time, there is a table `H` that contains the history of the table `V` at that point in time.
The table `H` has columns `from` and `to` that define the time interval for which the row is valid.
The table `H` also contains the details from table `V` at that point in time.

```{r}
H1 <- tibble(from = 1L, to = NA_integer_, V1)
H2 <- tibble(from = 1:2, to = NA_integer_, V2)
H3 <- tibble(
  from = 1:3,
  to = c(3L, NA_integer_, NA_integer_),
  bind_rows(V1[1, ], V3[2:1, ])
)
H4 <- tibble(
  from = 1:3,
  to = c(3L, NA_integer_, 4L),
  bind_rows(V1[1, ], V3[2:1, ])
)
```

With that, we can define a function `at_time_simple()` that takes a history table and a point in time, and returns the observation table at that point in time.

```{r}
at_time_simple <- function(V, time) {
  V |>
    filter(coalesce(from <= !!time, TRUE), coalesce(to > !!time, TRUE)) |>
    select(-from, -to) |>
    arrange(id)
}

H1 |>
  at_time_simple(1) |>
  waldo::compare(V1)

H2 |>
  at_time_simple(2) |>
  waldo::compare(V2)

H2 |>
  at_time_simple(1) |>
  waldo::compare(V1)

H3 |>
  at_time_simple(3) |>
  waldo::compare(V3)

H3 |>
  at_time_simple(2) |>
  waldo::compare(V2)

H3 |>
  at_time_simple(1) |>
  waldo::compare(V1)

H4 |>
  at_time_simple(4) |>
  waldo::compare(V4)

H4 |>
  at_time_simple(3) |>
  waldo::compare(V3)

H4 |>
  at_time_simple(2) |>
  waldo::compare(V2)

H4 |>
  at_time_simple(1) |>
  waldo::compare(V1)
```

## Decomposition

The history tables can be decomposed into two tables: `O` (observation) and `D` (difference).
The observation table contains the details from the history table at the point in time, and is identical to the data at that point in time, save for the `from` and `to` columns.
The difference table contains the changes that happened compared to the prior point in time.

Because we want to avoid storing the same data multiple times, we omit rows in a difference table that are identical to rows found in previous difference tables.

```{r}
O1 <- H1
O1 |>
  select(-from, -to) |>
  waldo::compare(V1)

D1 <- H2[0, ]

O2 <- H2
O2 |>
  select(-from, -to) |>
  waldo::compare(V2)

D2 <- H3[1, ]

O3 <- H3[3:2, ]
O3 |>
  select(-from, -to) |>
  waldo::compare(V3)

# This does not contain H4[1, ], on purpose:
D3 <- H4[3, ]

O4 <- H4[2, ]

O4 |>
  select(-from, -to) |>
  waldo::compare(V4)
```

Binding the an observed table and past history tables give a superset of the history table at that point in time.

```{r}
bind_rows(O1) |>
  distinct(id, .keep_all = TRUE) |>
  arrange(from, id) |>
  waldo::compare(H1)

bind_rows(O2, D1) |>
  distinct(id, .keep_all = TRUE) |>
  arrange(from, id) |>
  waldo::compare(H2)

bind_rows(O3, D2, D1) |>
  distinct(id, .keep_all = TRUE) |>
  arrange(from, id) |>
  waldo::compare(H3)

bind_rows(O4, D3, D2, D1) |>
  distinct(id, .keep_all = TRUE) |>
  arrange(from, id) |>
  waldo::compare(H4)
```

This allows constructing a function `at_time()` that takes an observation table and a difference table, and returns the observation table at that point in time.
The only additional step is to keep the first row for each id.
If the order plays a role, the rows can be sorted by `id`.

```{r}
at_time <- function(V, time) {
  V |>
    filter(coalesce(from <= !!time, TRUE), coalesce(to > !!time, TRUE)) |>
    # Keep the first row for each id
    distinct(id, .keep_all = TRUE) |>
    select(-from, -to) |>
    arrange(id)
}

O4 |>
  at_time(4)

bind_rows(O4, D3) |>
  at_time(3)

bind_rows(O4, D3, D2) |>
  at_time(2)

bind_rows(O4, D3, D2, D1) |>
  at_time(1)
```

## Generalization

Because observation and difference tables are a superset of history tables, combining, e.g., one observation table and two difference tables allows reconstructing the original data for three points in time in the past.

```{r}
bind_rows(O4, D3, D2, D1) |>
  at_time(1) |>
  waldo::compare(V1)

bind_rows(O4, D3, D2, D1) |>
  at_time(2) |>
  waldo::compare(V2)

bind_rows(O4, D3, D2) |>
  at_time(2) |>
  waldo::compare(V2)

bind_rows(O4, D3, D2) |>
  at_time(3) |>
  waldo::compare(V3)

bind_rows(O4, D3) |>
  at_time(3) |>
  waldo::compare(V3)

bind_rows(O4, D3) |>
  at_time(4) |>
  waldo::compare(V4)
```

## Updating observation and difference tables

How to construct `O4` and `D3` from `O3`, `D2`, `D1`, and `V4`?
Same question for constructing `O3` and `D2` from `O2`, `D1`, and `V3`?
Or for constructing `O2` and `D1` from `O1` and `V2`?
Or for the initialization, constructing `O1` from `V1`?

We know that we can reconstruct the history table from the observation and difference tables.
This then boils down to the question of how to construct `O4` and `D3` from `O3`, `H3`, and `V4`.

```{r}
O4
D3
O3
H3
V4
```

We know how to extract `V3` from `O3`:

```{r}
O3 |>
  at_time_simple(3) |>
  waldo::compare(V3)
```

We then can compute the new or updated, and deleted rows.
We also define `V0` and `O0` as the empty tables.

```{r}
V0 <- V1[0, ]
O0 <- O1[0, ]

compute_diff <- function(old, new) {
  # Contains both new and updated rows
  P <-
    new |>
    anti_join(old, by = names(new))

  # The id values of the deleted rows
  M <-
    old |>
    anti_join(new, by = "id") |>
    select(id)

  PM <-
    P |>
    select(id) |>
    bind_rows(M)

  list(P = P, M = M, PM = PM)
}

X4 <- compute_diff(V3, V4)
X4

X3 <- compute_diff(V2, V3)
X3

X2 <- compute_diff(V1, V2)
X2

X1 <- compute_diff(V0, V1)
X1
```

The observation table is the same as the new table with `from` and `to` set to the relevant points in time.
For new and updated rows, `from` is set to the current point in time; otherwise, the point in time from the old observation table is used.
Deleted rows must be removed from the observation table.

```{r}
X4$P |>
  mutate(from = 4L, to = NA_integer_, .before = 1) |>
  bind_rows(O3) |>
  distinct(id, .keep_all = TRUE) |>
  anti_join(X4$M, by = "id") |>
  arrange(id) |>
  waldo::compare(O4)

X3$P |>
  mutate(from = 3L, to = NA_integer_, .before = 1) |>
  bind_rows(O2) |>
  distinct(id, .keep_all = TRUE) |>
  anti_join(X3$M, by = "id") |>
  arrange(id) |>
  waldo::compare(O3)

X2$P |>
  mutate(from = 2L, to = NA_integer_, .before = 1) |>
  bind_rows(O1) |>
  distinct(id, .keep_all = TRUE) |>
  anti_join(X2$M, by = "id") |>
  arrange(id) |>
  waldo::compare(O2)

X1$P |>
  mutate(from = 1L, to = NA_integer_, .before = 1) |>
  bind_rows(O0) |>
  distinct(id, .keep_all = TRUE) |>
  anti_join(X1$M, by = "id") |>
  arrange(id) |>
  waldo::compare(O1)
```

The new difference table is the history table with the changed rows and `to` set to the current point in time.

```{r}
H3 |>
  semi_join(X4$PM, by = "id") |>
  filter(.by = id, row_number(from) == n()) |>
  mutate(to = 4L) |>
  waldo::compare(D3)

H2 |>
  semi_join(X3$PM, by = "id") |>
  filter(.by = id, row_number(from) == n()) |>
  mutate(to = 3L) |>
  waldo::compare(D2)

H1 |>
  semi_join(X2$PM, by = "id") |>
  filter(.by = id, row_number(from) == n()) |>
  mutate(to = 2L) |>
  waldo::compare(D1)
```

The first observation table is the same as the first table with `from` set to the first point in time and `to` set to missing.

```{r}
V1 |>
  mutate(from = 1L, to = NA_integer_, .before = 1) |>
  waldo::compare(O1)
```

This defines a process for efficiently maintaining the observation and difference tables as new data arrives.
